Decoupled Fallback: A Portable Single-Pass GPU Scan

### 2025-02-07 paper.tex
@b0nes164 just making some comments on these nice results

\section{Evaluation}
\subsection{Experimental Setup}
We configure workgroups with a size $w = 256$ threads, where each thread processes $p = 4$ vectors of size 4, resulting in 16 elements per thread and a total work tile size of $t = 4096$. Through testing, we find a maximum spin count of 4 to effectively balance responsiveness to blocking while minimizing false-positive fallbacks. As the size of the local spine is inversely proportional to the subgroup size, we allocate sufficient shared memory to accommodate the WebGPU standard's minimum subgroup size of 4. Due to our overall thriftiness with shared memory, total usage remains modest at only $\sim$1 KB. For all tests, we use an input size of $2^{25}$ and an inclusive prefix sum. Our test hardware, listed POSITIONAL-RELATION, spans a wide range of FPG and performance capabilities. We empirically find that the Google Chrome native implementation of WebGPU (Dawn) provides the best performance and use it for all testing, except on the Arm-Mali GPU, which requires Android native, necessitating the use of WGPU.\footnote{To keep the perforamance as close to Dawn as possible, we translate our WGSL code with Dawn's Tint transpiler, and provide the raw SPIR-V to WGPU.}
\begin{table}[t]
  \centering
  \small
  \begin{tabular*}{\linewidth}{@{\extracolsep{\fill}} l l c r r}
    \toprule
    Vendor & Device Name & FPG & Occupancy & Subgroup Size \\  
    \midrule
    Intel  & HD Graphics 620       & \textbf{Yes} & 9   & 8 - 32  \\  
    ARM    & Mali-G78 MP20          & \textbf{Yes} & 40  & 16  \\  
    Apple  & M1 Pro Max             & \textbf{No}  & 64  & 32  \\  
    NVIDIA & GeForce RTX 2080 Super & \textbf{Yes} & 192 & 32  \\  
    AMD    & Radeon RX 7900 XT      & \textbf{Yes} & 420 & 32 - 64  \\  
    \bottomrule
  \end{tabular*}
  \caption{Hardware Specifications of Testing Devices}
  \label{tab:hardware}
\end{table}

\subsection{Performance Comparison}
To evaluate the performance of \emph{Decoupled Fallback} (\emph{CSDLDF}), we compare it against two key benchmarks: \emph{Memcopy}\textdagger{~\ref{sec:memcopy}} and \emph{Reduce then Scan}\textdagger{~\ref{sec:rts}} (\emph{RTS}). \emph{RTS} serves as a baseline to demonstrate the performance gains of our approach over the previous FPG-constrained technique. \emph{Memcopy} is included as a point of reference, as it shares the same $2n$ global memory movement as \emph{Decoupled Fallback}, without any compute overhead and thus is guaranteed to be memory bandwidth bound.

\begin{table*}
  \centering
  \begin{tabular*}{\textwidth}{@{\extracolsep{\fill}} l r r r r r}
    \toprule
    Metric & Intel HD620 & ARM Mali-G78 MP20 & Apple M1 Pro Max & Nvidia 2080 Super & AMD 7900 XT \\  
    \midrule
    H-Mean \emph{Memcopy} (32-bit ele/sec)  & 1.498e9   & 106.663e9 & 54.203e9  & 54.203e9  & 71.190e9  \\
    H-Mean \emph{CSDLDF} (32-bit ele/sec)   & 1.487e9   & 112.430e9 & 53.256e9  & 53.256e9  & 95.737e9  \\  
    H-Mean \emph{RTS} (32-bit ele/sec)      & 1.060e9   & 82.302e9  & 35.709e9  & 35.709e9  & 62.868e9  \\  
    \emph{CSDLDF}/\emph{Memcopy} (\%)       & 99.2\%    & 105.4\%   & 98.3\%    & 98.3\%    & 134.5\%   \\  
    \emph{CSDLDF}/\emph{RTS} (\%)          & 140.2\%   & 136.6\%   & 149.1\%   & 149.1\%   & 152.3\%   \\  
    Avg Spins per Wg    & 0.306    & 1.380    & 0.418    & 0.418    & 0.133    \\  
    Avg Lookback Length per Wg  & 1.725    & 3.510    & 13.139   & 13.139   & 20.867   \\  
    Avg Total Fallbacks Initiated      & 260.364   & 2202.045  & 88.384   & 88.384   & 10.127   \\  
    Avg Total Successful Insertions    & 6.084     & 57.875    & 1.205    & 1.205    & 1.001    \\
    \bottomrule
  \end{tabular*}
  \caption{MALI ABSOLUTE SPEED WRONG!, \emph{Decoupled Fallback} performance, 32-bit Inclusive Prefix Sum, Input Size $\mathbf{2^{25}}$}
  \label{tab:results}
\end{table*}

As shown in Table~\ref{tab:results}, with the exception of the ARM Mali, \emph{CSDLDF} achieves at least a $\sim$40\% speedup over \emph{RTS} across all tested devices. Since \emph{Memcopy} is purely memory bandwidth bound, \emph{CSDLDF} should ideally match its throughput. This holds across most devices, except for the AMD 7900 XT, where \emph{CSDLDF} slightly outperforms \emph{Memcopy}, suggesting improved memory structuring due to compute operations. On the ARM Mali, \emph{CSDLDF} appears to underperform, achieving only a 36\% speedup instead of the expected 50\%. However, as its performance is nearly identical to \emph{Memcopy}, this indicates that the implementation is memory bandwidth bound, limiting further gains.

@@

where:
\begin{itemize}
    \item \texttt{<TestType>} specifies the test to run. Valid options are:
    \begin{itemize}
        \item \texttt{csdl} – Runs the \emph{Decoupled Lookback} test. \textbf{Warning}, this \emph{will} crash without FPG. 
        \item \texttt{csdldf} – Runs the \emph{Decoupled Fallback} test.
        \item \texttt{full} – Runs a full set of tests, including multiple \emph{Decoupled Fallback} simulations.
    \end{itemize}
    \item \texttt{record} (optional) enables recording the test results to a CSV file.
    \item \texttt{deviceName} (optional) appends the device name to recorded test results.
\end{itemize}

>>>>

\section{Evaluation}
\subsection{Experimental Setup}
We configure workgroups with a size $w = 256$ threads, where each thread processes $p = 4$ vectors of size 4, resulting in 16 elements per thread and a total work tile size of $t = 4096$. Through testing, \john{across all devices (?)} we find a maximum spin count of 4 effectively balances responsiveness to blocking while minimizing false-positive fallbacks. As the size of the local spine is inversely proportional to the subgroup size, we allocate sufficient shared memory to accommodate the WebGPU standard's minimum subgroup size of 4. Due to our overall thriftiness with shared memory, total usage remains modest at only $\sim$1 KB\@. For all tests, we use an input size of $2^{25}$ \john{It would be worthwhile, I think, to run a test against CUB where the x axis is problem size and the y axis is throughput. A reviewer might think we're hiding something if we only report results for one (large) input size. I imagine that our performance will look similar to CUB's at all input sizes and that is the result we want.} and an inclusive prefix sum. Our test hardware, listed POSITIONAL-RELATION, spans a wide range of FPG and performance capabilities. We empirically find that the Google Chrome native implementation of WebGPU (Dawn) provides the best performance and use it for all testing, except on the Arm-Mali GPU, which requires Android native, necessitating the use of WGPU\@.\footnote{To keep the perforamance as close to Dawn as possible, we translate our WGSL code with Dawn's Tint transpiler, and provide the raw SPIR-V to WGPU\@.}
\begin{table}
  \centering
  \small
  \begin{tabular*}{\linewidth}{@{\extracolsep{\fill}} l l c r r}
    \toprule
    Vendor & Device Name & FPG & Occupancy & Subgroup Size \\
    \midrule
    Intel  & HD Graphics 620       & \textbf{Yes} & 9   & 8--32  \\
    ARM    & Mali-G78 MP20          & \textbf{Yes} & 40  & 16  \\
    Apple  & M1 Pro Max             & \textbf{No}  & 64  & 32  \\
    NVIDIA & GeForce RTX 2080 Super & \textbf{Yes} & 192 & 32  \\
    AMD    & Radeon RX 7900 XT      & \textbf{Yes} & 420 & 32--64  \\
    \bottomrule
  \end{tabular*}
  \caption{Hardware Specifications of Testing Devices.}
  \label{tab:hardware}
\end{table}

\subsection{Performance Comparison}
To evaluate the performance of \emph{Decoupled Fallback} (\emph{CSDLDF}), we compare it against two key benchmarks: \emph{Memcopy}\textdagger{~\ref{sec:memcopy}} and \emph{Reduce then Scan}\textdagger{~\ref{sec:rts}} (\emph{RTS}). \emph{RTS} serves as a baseline to demonstrate the performance gains of our approach over the previous FPG-constrained technique. \emph{Memcopy} is included as a point of reference, as it shares the same $2n$ global memory movement as \emph{Decoupled Fallback}, without any compute overhead and thus is guaranteed to be memory bandwidth bound. \john{Nit: I would use \texttt{memcpy} as the name of this benchmark to match how it's actually called in C etc.}

\begin{table*}
  \centering
  \begin{tabular*}{\textwidth}{@{\extracolsep{\fill}} l r r r r r}
    \toprule
    Metric & Intel HD620 & ARM Mali-G78 MP20 & Apple M1 Pro Max & Nvidia 2080 Super & AMD 7900 XT \\
    \midrule
    H-Mean \emph{Memcopy} (32-bit ele/sec)  & 1.498e9   & 106.663e9 & 54.203e9  & 54.203e9  & 71.190e9  \\
    H-Mean \emph{CSDLDF} (32-bit ele/sec)   & 1.487e9   & 112.430e9 & 53.256e9  & 53.256e9  & 95.737e9  \\
    H-Mean \emph{RTS} (32-bit ele/sec)      & 1.060e9   & 82.302e9  & 35.709e9  & 35.709e9  & 62.868e9  \\
    \emph{CSDLDF}/\emph{Memcopy} (\%)       & 99.2\%    & 105.4\%   & 98.3\%    & 98.3\%    & 134.5\%   \\
    \emph{CSDLDF}/\emph{RTS} (\%)          & 140.2\%   & 136.6\%   & 149.1\%   & 149.1\%   & 152.3\%   \\
    Avg Spins per Wg    & 0.306    & 1.380    & 0.418    & 0.418    & 0.133    \\
    Avg Lookback Length per Wg  & 1.725    & 3.510    & 13.139   & 13.139   & 20.867   \\
    Avg Total Fallbacks Initiated      & 260.364   & 2202.045  & 88.384   & 88.384   & 10.127   \\
    Avg Total Successful Insertions    & 6.084     & 57.875    & 1.205    & 1.205    & 1.001    \\
    \bottomrule
  \end{tabular*}
  \caption{MALI ABSOLUTE SPEED WRONG!, \emph{Decoupled Fallback} performance, 32-bit Inclusive Prefix Sum, Input Size $\mathbf{2^{25}}$. \john{I would label metrics as ``H-Mean \emph{Memcopy} (32-bit ele/sec $\cdot 10^9$)'' and then leave the exponents out of the main table.}\john{Separately, h-mean is the right summary statistic \emph{if} the distribution is Gaussian, which it almost certainly is not. Like most measurements of this type, it's much more likely log-normal. I am influenced here by \url{https://lemire.me/blog/2023/04/06/are-your-memory-bound-benchmarking-timings-normally-distributed/}.}\john{I note Apple and NVIDIA timings are identical, but figure you know that.}}
  \label{tab:results}
\end{table*}

As shown in Table~\ref{tab:results}, with the exception of the ARM Mali, \emph{CSDLDF} achieves at least a $\sim$40\% speedup over \emph{RTS} across all tested devices. Since \emph{Memcopy} is purely memory bandwidth bound, \emph{CSDLDF} should ideally match its throughput. This holds across most devices, except for the AMD 7900 XT, where \emph{CSDLDF} slightly outperforms \emph{Memcopy}, suggesting improved memory structuring due to compute operations. \john{I have no idea what this previous phrase means.} On the ARM Mali, \emph{CSDLDF} appears to underperform, achieving only a 36\% speedup instead of the expected 50\%. However, as its performance is nearly identical to \emph{Memcopy}, this indicates that the implementation is memory-bandwidth-bound, limiting further gains.

@@

where:
\begin{itemize}
    \item \texttt{<TestType>} specifies the test to run. Valid options are:
    \begin{itemize}
        \item \texttt{csdl} – Runs the \emph{Decoupled Lookback} test. \textbf{Warning}, this \emph{will} crash without FPG\@.
        \item \texttt{csdldf} – Runs the \emph{Decoupled Fallback} test.
        \item \texttt{full} – Runs a full set of tests, including multiple \emph{Decoupled Fallback} simulations.
    \end{itemize}
    \item \texttt{record} (optional) enables recording the test results to a CSV file.
    \item \texttt{deviceName} (optional) appends the device name to recorded test results.
\end{itemize}
